---
title: "TidyTuesday: {{week_title}}"
description: "{{description}}"
date: "{{date}}"
author: "chokotto"
categories:
  - TidyTuesday
  - R
  - Finance
image: "thumbnail.svg"
engine: knitr
code-fold: true
---

## Overview

{{overview}}

## Dataset

```{r}
#| label: load-packages
#| message: false
#| warning: false

library(tidyverse)
library(ggplot2)
library(scales)
library(glue)
```

```{r}
#| label: load-data
#| message: false

# Load this week's data
# tuesdata <- tidytuesdayR::tt_load('{{date}}')
```

## Exploratory Analysis

{{analysis}}

## Visualization

```{r}
#| label: main-viz
#| fig-width: 12
#| fig-height: 7
#| warning: false

# Main visualization code here
```

## Key Findings

{{findings}}

---
_This post is part of the [TidyTuesday](https://github.com/rfordatascience/tidytuesday) weekly data visualization project._

:::{.callout-caution collapse="false" appearance="minimal" icon="false"}
## Disclaimer

::: {style="font-size: 0.85em; color: #64748b; line-height: 1.6;"}
This analysis is for educational and practice purposes only. Data visualizations and interpretations are based on the provided dataset and may not represent complete or current information.
:::
:::
